Dialect: A Compact Line Notation Based on SMILES

### 2022-03-24 manuscript.tex
Uses smart quotes. closes #100

Despite its many applications, SMILES remains a language with an incomplete specification. Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of "aromatic" features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.

@@

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2008}. The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. ("Daylight"), recapitulates previously-published material~\cite{daylightTheory}. The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.

In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}. OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed. Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.

In 2019 IUPAC announced the SMILES+ initiative~\cite{smilesPlusSpecification}. Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project. Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted~\cite{smilesPlusGitHub}.

@@

The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts. The FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}. Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.

This situation has persisted long enough that the notion of a SMILES "standard" may not currently be feasible. A dozen or more slightly different implementations are now in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language "drift" while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of written documentation. The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use~\cite{daylightToolkit}. For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned~\cite{depictArchive}. Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Dialect's system of representation is based on the \textit{molecular graph} concept~\cite{balaban:1985}. A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds"). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and Molfile~\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

A Dialect molecule consists of a graph with zero or more nodes and zero or more edges. The "empty molecule", devoid of atoms and bonds, is allowed. Dialect imposes no upper bound on the number of atoms or bonds. However, practical limitations related to memory, storage, and CPU time will likely limit the size of molecules in practice.

@@

Table \ref{table:atomic-state} summarizes atomic attributes. Not all combinations of values ("atomic state") are valid. Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created.

@@

Dialect supports two mutually-exclusive modes of hydrogen suppression. The first is \textit{virtual hydrogen}. A virtual hydrogen is one whose node and connecting bond have been replaced by an integer tally attribute on the parent atom. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four. When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive. The process of transforming one or more hydrogen nodes into a virtual hydrogen is called "virtualization." The opposite process, transforming one or more virtual hydrogens into a node, is called "reification."

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME occurs when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization. Two molecular graphs are equal if they are isomorphic and a 1:1 mapping exists between the atoms and bonds in each graph. Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules. DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} ("selection"). For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}. Atoms having other elements (or \ttt{None}) must not be selected. Only selected atoms belong to the DS.

@@

A filled DS can be emptied through a two-step process of \textit{deselection}. First, a perfect matching over the DS is found. Next, each matched edge is replaced by a double bond. Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds. A popular algorithm for matching, the Edmunds "blossom algorithm," has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}. Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.

@@

The opposite operation can be accomplished with a \textit{selection algorithm}. A selection algorithm selects two or more atoms, thereby adding them to the DS. The only requirement for a selection algorithm is that the resulting DS must have a perfect matching. Depending on the application, other criteria may be applied. For example, a selection algorithm can restrict candidate atoms to those found in cycles. Electron-counting techniques can also be introduced to approximate the chemical concept of "aromaticity." A double bond between two selected atoms may or may not be elided, depending on the application.

@@

To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute ("index"). The value of an atom's index equals zero for a molecular's graph's first atom. The index for each subsequently-added atom equals the order (node count) of the graph prior to the addition. For example, the index for the second atom is one, the index for the third atom is two, and so on. The maximum value of an atomic index is 2\textsuperscript{32} - 1.

@@

To support this system, each bond carries a nullable \ttt{direction} attribute ("direction"). When present, this attribute may assume one of the two values \ttt{Up} or \ttt{Down}. These values refer to a geometrical model in which the two terminals of a double bond and their neighbors are assigned local relative coordinates in a plane. The double bond terminals are placed on the x-axis such that the terminal with the lowest \ttt{index} attribute appears to the left of the terminal with the higher atomic \ttt{index} attribute. These terminals are designated "left terminal" and "right terminal," respectively. Each neighbor of a terminal is then assigned a relative coordinate based on the \ttt{index} attribute and the direction of its PPB bond.

@@

The third major element of molecular representation is \textit{configuration}. Configuration is the three-dimensional arrangement of neighbors about an atom. Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertexes of a tetrahedron. Here, "substituent" means an atomic neighbor; lone electron pairs are not considered substituents. One of the four substituents may be a virtual hydrogen.

A configuration is comprised of two components: an ordering of bonds to a central atom; and a \textit{configurational descriptor}. A configurational descriptor is a template for the relative three-dimensional positioning of neighbors about a central atom. Dialect supports two configurational descriptors: \ttt{Left} and \ttt{Right}. The context for the terms "left" and "right" will be explained shortly.

@@

Not all tetracoordinate atoms will be \textit{stereocenters}. A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel~\cite{mislow:1984}. In the context of Dialect, a stereocenter must be tetracoordinate. Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

@@

A Dialect string ("string") is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}. The internal structure of a string reflects a depth-first traversal of the corresponding molecular graphs. As such, the syntax supports branches, cycles, and disconnected components.

@@

Dialect's formal grammar is presented as a series of \textit{production rules} (aka "productions"). A production rule defines a transformation allowed under the grammar. These transformations collectively define the set of valid Dialect strings. Production rules can be used in the forward direction, when writing a string, or in the reverse direction, when reading a string.

A production rule is composed of two kinds of elements: \textit{terminals} and \textit{non-terminals}. A terminal is a character literal (e.g., "A"). A non-terminal is a reference to another production rule. This reference occurs through a name, which appears to the left of a separator (\ttt{::=}) in a production rule. To the right of a separator appear the allowed terminals and non-terminals for the rule.

Consider a hypothetical language composed of variable-length sequences of the lowercase letter "a". Such a language could be cast as the following two production rules:

@@

The quantifier (\ttt{*}) indicates that a text in this language consists of a sequence of zero or more instances of the production \ttt{<a>}. This production in turn is defined as the lowercase letter "a". Therefore, valid texts in this language include the empty string, \ttt{a}, \ttt{aa}, and \ttt{aaaaaa} to name a few.

@@

The first atomic non-terminal, \ttt{<star>}, can assume one terminal value, the asterisk character (\ttt{*}). This "star atom" represents an atom in which every attribute assumes its default value. The implicit hydrogen count of a star atom is zero.

@@

The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: "B"; "C"; "N"; "O"; "P"; "S"; "F"; "Cl"; "Br"; and "I". An atom encoded in this way ("shortcut atom") assigns the corresponding symbol to the \ttt{element} attribute. All other attributes retain their default values. The implicit hydrogen count of a shortcut atom is determined by the algorithm previously given.

@@

The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: "c"; "n"; "o"; "p"; and "s". An atom encoded in this way ("selected shortcut atom") assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`. All other atomic attributes retain their default values. The implicit hydrogen count of a selected shortcut atom is determined as described previously.

@@

The fourth and most complex atomic production rule is \ttt{<bracket>} ("bracket atom"). A bracket atom can be used to set any atomic attribute. A bracket atom must be used when the `element` attribute of an atom makes it ineligible for implicit hydrogen counting. Attributes not set within the \ttt{<bracket>} production rule will leave the corresponding atomic values in their default states.

@@

The value of a bracket atom's \ttt{isotope} attribute is determined by the optional \ttt{<isotope>} non-terminal. It consists of between one and three digits encoding the integers 0-999. Leading zeros are disregarded when assigning the value of the \ttt{isotope} attribute. In other words, "007" and "7" are considered equivalent expressions of the value 7.

@@

The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule ("symbol"). Three non-terminal options are available: \ttt{<star>}; \ttt{<element>}; and \ttt{<selection>}.

@@

The \ttt{stereodescriptor} attribute of a bracket atom is determined by the \ttt{<stereodescriptor>} non-terminal. Allowed values are "@" and "@@", corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.

@@

The \ttt{hydrogens} attribute of a bracket atom is controlled by the \\ \ttt{<virtual{\_}hydrogen>} non-terminal. This non-terminal is comprised of the terminal "H" followed by an optional \ttt{<digit>} non-terminal. A digit appearing after the "H" terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count. A digit of 0 (i.e., "H0") sets the \ttt{hydrogens} attribute to zero. If no digit is present, \ttt{hydrogens} is set to one.

@@

The \ttt{<charge>} production rule sets the \ttt{charge} attribute of a bracket atom. This non-terminal begins with either the plus or minus terminals ("+", "-", respectively) and ends with an optional \ttt{<digit>} non-terminal. A missing digit causes the atomic \ttt{charge} attribute to be set to one. In other words, "+" sets the \ttt{charge} attribute to one and "-" sets it to minus one. A digit of zero (e.g., "+0" or "-0") sets the \ttt{charge} attribute to zero.

@@

Should an optional bracket non-terminal not be present, the atom must retain its corresponding default value. For example, the bracket "[C@H+]" lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}. Similarly, the bracket "[13CH+]" lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.

@@

Five terminals are available ("-", "=", "\#", "/", and "\textbackslash"). The first three ("-", "=", and "\#") set the \ttt{order} attribute of a bond to 1; 2; or 3, respectively. The last two terminals, "/" and "\textbackslash", set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.

@@

An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal ("sequence"). A sequence starts with a required \ttt{<atom>} non-terminal. If no allowed non-terminals follow, the corresponding atom will have no children. Allowed non-terminals are chosen from the list: \ttt{<union>}; \ttt{<branch>}; and \ttt{<sequence>}.

@@

The \ttt{<cut>} non-terminal ("cut") can takes two forms. A digit non-terminal can be used, enabling single-digit cut indexes. Double- or single-digit cut indexes, supporting the values zero through 99 inclusive, are available by prepending the percent character (\ttt{\%}).

@@

An alternative to \ttt{<union>} within a sequence is the \ttt{<branch>} non-terminal ("branch"). Like union, branch joins a parent and child node through a bond. Wrapped by opening and closing parenthesis terminals ("(" and ")", respectively), branch encodes a sequence that may or may not be attached to its parent. Attachment occurs if the \ttt{<bond>} non-terminal is included. Alternatively, the subgraph will be detached if the \ttt{<detachment>} non-terminal appears. If neither \ttt{<bond>} nor \ttt{<detachment>} non-terminals are detected between the parentheses but a sequence is, then bonding of parent and child occurs through an elided bond.

@@

The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal ("split"). A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} non-terminals. The `<detachment>` non-terminal serves the same purpose as it does within a branch: the prevent the bonding of the parent atom with a child.

@@

The presence of a branch adds some nuances over a union. The leading open parenthesis terminal ("(") signifies that the current root atom will later be re-exposed. This operation can be supported by a stack. At the start of a branch, the current root is pushed to the stack. At the end of the branch, the stack is popped and its top value is assigned as the new root atom.

@@

Readers must not assume that detachment (the period terminal, ".") implies the presence of disconnected components. This assumption is most likely to arise in the context of ad-hoc parsers using regular expressions, string matching, and the like. For example, the single molecule propane can be encoded using the string "C1C.C1".

The partial parity bond model is unusual in that it spreads conformational information over three or more bonds. Regardless, readers must report as an error any string containing an isolated PPB (e.g., "C/C"). Beyond this check, generating (\textit{E}) and (\textit{Z}) stereodescriptors from PPBs is not a straightforward process. Performing such a conversion requires a method capable of translating PPB into a localized bond descriptor. One approach is to first consider the parity of each double bond terminal individually. A separate step would then relate the individual terminal parities to a double bond conformation parity. This intermediate parity can then be used to perform the final conversion.

@@

Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider "organic molecules." This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.

>>>>

Despite its many applications, SMILES remains a language with an incomplete specification. Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of ``aromatic'' features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.

@@

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2008}. The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. (``Daylight''), recapitulates previously-published material~\cite{daylightTheory}. The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.

In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}. OpenSMILES was conceived as ``a non-proprietary specification for the SMILES language.'' Many of the points left open by previous SMILES documentation efforts were addressed. Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of ``standard form.'' Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.

In 2019 IUPAC announced the SMILES+ initiative~\cite{smilesPlusSpecification}. Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to ``establish a formalized recommended up-to-date specification of the SMILES format.'' SMILES+ took as its starting point the documentation produced by the OpenSMILES project. Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted~\cite{smilesPlusGitHub}.

@@

The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts. The FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}. Interoperability in particular requires that ``(meta)data'' use a formal, accessible, shared, and broadly applicable language for knowledge representation.'' Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.

This situation has persisted long enough that the notion of a SMILES standard may not currently be feasible. A dozen or more slightly different implementations are now in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language drift while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of written documentation. The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use~\cite{daylightToolkit}. For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned~\cite{depictArchive}. Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Dialect's system of representation is based on the \textit{molecular graph} concept~\cite{balaban:1985}. A graph is a data structure comprised of a set of nodes (which map to atoms) and a set of pairwise relationships between them called edges (which map to bonds). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and Molfile~\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

A Dialect molecule consists of a graph with zero or more nodes and zero or more edges. The empty molecule, devoid of atoms and bonds, is allowed. Dialect imposes no upper bound on the number of atoms or bonds. However, practical limitations related to memory, storage, and CPU time will likely limit the size of molecules in practice.

@@

Table \ref{table:atomic-state} summarizes atomic attributes. Not all combinations of values (``atomic state'') are valid. Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created.

@@

Dialect supports two mutually-exclusive modes of hydrogen suppression. The first is \textit{virtual hydrogen}. A virtual hydrogen is one whose node and connecting bond have been replaced by an integer tally attribute on the parent atom. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four. When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive. The process of transforming one or more hydrogen nodes into a virtual hydrogen is called ``virtualization.'' The opposite process, transforming one or more virtual hydrogens into a node, is called ``reification.''

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME occurs when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization. Two molecular graphs are equal if they are isomorphic and a 1:1 mapping exists between the atoms and bonds in each graph. Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules. DIME is similar to ``resonance'' or ``aromaticity,'' but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} (``selection''). For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}. Atoms having other elements (or \ttt{None}) must not be selected. Only selected atoms belong to the DS.

@@

A filled DS can be emptied through a two-step process of \textit{deselection}. First, a perfect matching over the DS is found. Next, each matched edge is replaced by a double bond. Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds. A popular algorithm for matching, the Edmunds ``blossom algorithm,'' has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}. Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.

@@

The opposite operation can be accomplished with a \textit{selection algorithm}. A selection algorithm selects two or more atoms, thereby adding them to the DS. The only requirement for a selection algorithm is that the resulting DS must have a perfect matching. Depending on the application, other criteria may be applied. For example, a selection algorithm can restrict candidate atoms to those found in cycles. Electron-counting techniques can also be introduced to approximate the chemical concept of ``aromaticity.'' A double bond between two selected atoms may or may not be elided, depending on the application.

@@

To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute (``index''). The value of an atom's index equals zero for a molecular's graph's first atom. The index for each subsequently-added atom equals the order (node count) of the graph prior to the addition. For example, the index for the second atom is one, the index for the third atom is two, and so on. The maximum value of an atomic index is 2\textsuperscript{32} - 1.

@@

To support this system, each bond carries a nullable \ttt{direction} attribute (``direction''). When present, this attribute may assume one of the two values \ttt{Up} or \ttt{Down}. These values refer to a geometrical model in which the two terminals of a double bond and their neighbors are assigned local relative coordinates in a plane. The double bond terminals are placed on the x-axis such that the terminal with the lowest \ttt{index} attribute appears to the left of the terminal with the higher atomic \ttt{index} attribute. These terminals are designated ``left terminal'' and ``right terminal,'' respectively. Each neighbor of a terminal is then assigned a relative coordinate based on the \ttt{index} attribute and the direction of its PPB bond.

@@

The third major element of molecular representation is \textit{configuration}. Configuration is the three-dimensional arrangement of neighbors about an atom. Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertexes of a tetrahedron. Here, ``substituent'' means an atomic neighbor; lone electron pairs are not considered substituents. One of the four substituents may be a virtual hydrogen.

A configuration is comprised of two components: an ordering of bonds to a central atom; and a \textit{configurational descriptor}. A configurational descriptor is a template for the relative three-dimensional positioning of neighbors about a central atom. Dialect supports two configurational descriptors: \ttt{Left} and \ttt{Right}. The context for the terms ``left'' and ``right'' will be explained shortly.

@@

Not all tetracoordinate atoms will be \textit{stereocenters}. A stereocenter is an atom whose ``ligand permutation produces stereoisomers,'' as defined by Mislow and Siegel~\cite{mislow:1984}. In the context of Dialect, a stereocenter must be tetracoordinate. Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

@@

A Dialect string (``string'') is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}. The internal structure of a string reflects a depth-first traversal of the corresponding molecular graphs. As such, the syntax supports branches, cycles, and disconnected components.

@@

Dialect's formal grammar is presented as a series of \textit{production rules} (aka ``productions''). A production rule defines a transformation allowed under the grammar. These transformations collectively define the set of valid Dialect strings. Production rules can be used in the forward direction, when writing a string, or in the reverse direction, when reading a string.

A production rule is composed of two kinds of elements: \textit{terminals} and \textit{non-terminals}. A terminal is a character literal (e.g., ``A''). A non-terminal is a reference to another production rule. This reference occurs through a name, which appears to the left of a separator (\ttt{::=}) in a production rule. To the right of a separator appear the allowed terminals and non-terminals for the rule.

Consider a hypothetical language composed of variable-length sequences of the lowercase letter ``a''. Such a language could be cast as the following two production rules:

@@

The quantifier (\ttt{*}) indicates that a text in this language consists of a sequence of zero or more instances of the production \ttt{<a>}. This production in turn is defined as a lowercase letter (`a`). Therefore, valid texts in this language include the empty string, \ttt{a}, \ttt{aa}, and \ttt{aaaaaa} to name a few.

@@

The first atomic non-terminal, \ttt{<star>}, can assume one terminal value, the asterisk character (\ttt{*}). This ``star atom'' represents an atom in which every attribute assumes its default value. The implicit hydrogen count of a star atom is zero.

@@

The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: ``B''; ``C''; ``N''; ``O''; ``P''; ``S''; ``F''; ``Cl''; ``Br''; and ``I.'' An atom encoded in this way (``shortcut atom'') assigns the corresponding symbol to the \ttt{element} attribute. All other attributes retain their default values. The implicit hydrogen count of a shortcut atom is determined by the algorithm previously given.

@@

The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: ``c''; ``n''; ``o''; ``p''; and ``s.'' An atom encoded in this way (``selected shortcut atom'') assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`. All other atomic attributes retain their default values. The implicit hydrogen count of a selected shortcut atom is determined as described previously.

@@

The fourth and most complex atomic production rule is \ttt{<bracket>} (``bracket atom''). A bracket atom can be used to set any atomic attribute. A bracket atom must be used when the `element` attribute of an atom makes it ineligible for implicit hydrogen counting. Attributes not set within the \ttt{<bracket>} production rule will leave the corresponding atomic values in their default states.

@@

The value of a bracket atom's \ttt{isotope} attribute is determined by the optional \ttt{<isotope>} non-terminal. It consists of between one and three digits encoding the integers 0-999. Leading zeros are disregarded when assigning the value of the \ttt{isotope} attribute. In other words, ``007'' and ``7'' are considered equivalent expressions of the value 7.

@@

The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule (``symbol''). Three non-terminal options are available: \ttt{<star>}; \ttt{<element>}; and \ttt{<selection>}.

@@

The \ttt{stereodescriptor} attribute of a bracket atom is determined by the \ttt{<stereodescriptor>} non-terminal. Allowed values are ``@'' and ``@@,'' corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.

@@

The \ttt{hydrogens} attribute of a bracket atom is controlled by the \\ \ttt{<virtual{\_}hydrogen>} non-terminal. This non-terminal is comprised of the terminal ``H'' followed by an optional \ttt{<digit>} non-terminal. A digit appearing after the ``H'' terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count. A digit of 0 (i.e., ``H0'') sets the \ttt{hydrogens} attribute to zero. If no digit is present, \ttt{hydrogens} is set to one.

@@

The \ttt{<charge>} production rule sets the \ttt{charge} attribute of a bracket atom. This non-terminal begins with either the plus or minus terminals (``+'' and ``-,'' respectively) and ends with an optional \ttt{<digit>} non-terminal. A missing digit causes the atomic \ttt{charge} attribute to be set to one. In other words, the plus terminal (``+'') sets the \ttt{charge} attribute to one and the minus terminal (``-'') sets it to minus one. A charge of zero (e.g., ``+0'' or ``-0'') sets the \ttt{charge} attribute to zero.

@@

Should an optional bracket non-terminal not be present, the atom must retain its corresponding default value. For example, the bracket ``[C@H+]'' lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}. Similarly, the bracket ``[13CH+]'' lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.

@@

Five terminals are available (``-,'' ``=,'' ``\#,'' ``/,'' and ``\textbackslash''). The first three (``-,'' ``=,'' and ``\#'') set the \ttt{order} attribute of a bond to one, two, or three, respectively. The last two terminals, ``/'' and ``\textbackslash,'' set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.

@@

An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal (``sequence''). A sequence starts with a required \ttt{<atom>} non-terminal. If no allowed non-terminals follow, the corresponding atom will have no children. Allowed non-terminals are chosen from the list: \ttt{<union>}; \ttt{<branch>}; and \ttt{<sequence>}.

@@

The \ttt{<cut>} non-terminal (``cut'') can takes two forms. A digit non-terminal can be used, enabling single-digit cut indexes. Double- or single-digit cut indexes, supporting the values zero through 99 inclusive, are available by prepending the percent character (\ttt{\%}).

@@

An alternative to \ttt{<union>} within a sequence is the \ttt{<branch>} non-terminal (``branch''). Like union, branch joins a parent and child node through a bond. Wrapped by opening and closing parenthesis terminals (``('' and ``),'' respectively), branch encodes a sequence that may or may not be attached to its parent. Attachment occurs if the \ttt{<bond>} non-terminal is included. Alternatively, the subgraph will be detached if the \ttt{<detachment>} non-terminal appears. If neither \ttt{<bond>} nor \ttt{<detachment>} non-terminals are detected between the parentheses but a sequence is, then bonding of parent and child occurs through an elided bond.

@@

The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal (``split''). A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} non-terminals. The `<detachment>` non-terminal serves the same purpose as it does within a branch: the prevent the bonding of the parent atom with a child.

@@

The presence of a branch adds some nuances over a union. The leading open parenthesis terminal (``('') signifies that the current root atom will later be re-exposed. This operation can be supported by a stack. At the start of a branch, the current root is pushed to the stack. At the end of the branch, the stack is popped and its top value is assigned as the new root atom.

@@

Readers must not assume that detachment (the period terminal, ``.'') implies the presence of disconnected components. This assumption is most likely to arise in the context of ad-hoc parsers using regular expressions, string matching, and the like. For example, the single molecule propane can be encoded using the string ``C1C.C1''.

The partial parity bond model is unusual in that it spreads conformational information over three or more bonds. Regardless, readers must report as an error any string containing an isolated PPB (e.g., ``C/C''). Beyond this check, generating (\textit{E}) and (\textit{Z}) stereodescriptors from PPBs is not a straightforward process. Performing such a conversion requires a method capable of translating PPB into a localized bond descriptor. One approach is to first consider the parity of each double bond terminal individually. A separate step would then relate the individual terminal parities to a double bond conformation parity. This intermediate parity can then be used to perform the final conversion.

@@

Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider ``organic molecules.'' This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.
